---
title: "2 comparing climate data"
author: "Matthew Ross"
date: "2024-04-17"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(RcppRoll)

gldas_dir_y= 'data/yampa/'

if(!file.exists(gldas_dir_y)){
  dir.create('data')
  dir.create(gldas_dir_y)
}
```

# Assignment

For this assignment we are going to compare climate data from a single
point versus a watershed averaged climate data. We will be working over
the Yampa river watershed, one of the last undammed watersheds in the
USA.

## Point-climate acquisition.

Using the 1_climate_data_yojoa.Rmd as an example download at least two
types of climate (wind/temp/etc...) GLDAS data for the Yampa River above
Elkhead Creek. This is the site of a USGS gage that has data dating back
to 2004.

```{r}

# Using GLDAS 2.1 for this assignment
yampa_gldas_mod = 'GLDAS_NOAH025_3H_v2.1'

site_info <- tibble(site_no = '09244490',
                    lat = 40.5180278,
                    long = -107.3997838,
                    name = 'Yampa_hayden')

start_date <- '2004-01-01'
end_date <- '2023-01-01'

# Downloading rain and temperature
yampa_params = c('Tair_f_inst', 'Rainf_f_tavg')
```

```{r}

# setting up function to download data:
yampa_make_www_2.1 = function(model, var){#, s_d, e_d, lat, lon) {
  s_d = start_date
  e_d = end_date
  paste0('https://hydro1.gesdisc.eosdis.nasa.gov/daac-bin/access/timeseries.cgi?variable=GLDAS2:', model, ':', var, '&startDate=', s_d, 'T00:00&endDate=', e_d, 'T21:00&location=GEOM:POINT(', lon, ',%20', lat, ')&type=asc2')
}


#make a list of all wwws for download
yampa_v2.1_www = unlist(map2(rep(yampa_gldas_mod, times = length(yampa_params)), yampa_params, yampa_make_www_2.1))


# Downloading the data:
for(w21 in 1:length(yampa_v2.1_www)) {
  download.file(url = yampa_v2.1_www[w21], destfile = file.path(gldas_dir_y, paste0(rep(yampa_gldas_mod, length(yampa_params))[w21], '_', yampa_params[w21], '.csv')))
}
```

```{r}
# loading in the data and collating:

#make list of files for all GLDAS data
yampafiles = list.files(gldas_dir_y)

formatGLDAS = function(file = files[1]){
  f = read.delim(file.path(gldas_dir_y, file), skip = 12, sep = '\t')
  colnames(f) = 'data'
  f = f %>%
    rownames_to_column('datetime') %>% 
    mutate(data = as.character(data),
           datetime = as.character(datetime),
           parameter = unlist(str_split(file, pattern = '_'))[5],
           version = unlist(str_split(file, pattern = '_'))[4])
  return(f)
}

# making df:
yampa_all_gldas = map_dfr(yampafiles, formatGLDAS) %>% 
  mutate(datetime_gmt = as.POSIXct(datetime, tz = 'Etc/GMT+0'),
         data = as.numeric(data)) %>%
  arrange(datetime_gmt)

# check that data looks good:
View(yampa_all_gldas)

# plotting to double check:
ggplot(yampa_all_gldas %>%
         sample_frac(0.05), aes(x = datetime_gmt, y = data)) +
  geom_point() +
  facet_grid(parameter ~ ., scales = 'free_y') +
  theme_bw()

```

```{r}
# making horizontal
yampa_all_gldas_h = all_gldas %>% 
  group_by(datetime, parameter) %>% 
  summarise(aggrate_data = mean(data)) %>% 
  pivot_wider(names_from = c('parameter'),
              values_from = 'aggrate_data')
```

## Watershed averaged climate data

Using climate engine, download the two same parameters but for the
watershed. The watershed is stored as a .geojson in the yampa folder.
Note, you likely need to convert the watershed to a ".shp" file before
getting the data from climate engine.

## Compare your climate data anyway you want

Make at least two plots comparing your point versus watershed-averaged
climate data.
