---
title: "2 comparing climate data"
author: "Matthew Ross"
date: "2024-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(RcppRoll)
library(sf)

gldas_dir = 'data/GLDAS/'

if(!file.exists(gldas_dir)){
  dir.create('data')
  dir.create(gldas_dir)
}
```


# Assignment


For this assignment we are going to compare climate data from a single point 
versus a watershed averaged climate data. We will be working over the Yampa river
watershed, one of the last undammed watersheds in the USA. 

## Point-climate acquisition. 

Using the 1_climate_data_yojoa.Rmd as an example download at least two types of climate (wind/temp/etc...) GLDAS data for the
Yampa River above Elkhead Creek. This is the site of a USGS gage that has data
dating back to 2004. 

```{r}
site_info <- tibble(site_no = '09244490',
                    lat = "40.5180278",
                    long = "-107.3997838",
                    name = 'Yampa_hayden')

#set parameters of function - these are found on the data rods website linked above
gldas_mod = "GLDAS_NOAH025_3H_v2.1"
params = c('Wind_f_inst', 'Rainf_f_tavg') # shortened param list to only two

#time period of interest - v 2.1 only because time range is 2004 - present
start_date_2.1 = '2004-01-01'
end_date_2.1 = '2023-04-18'

#function to make wwws to ping - 2.1 only
make_www_2.1 = function(model, var){ # s_d, e_d are inputs
  s_d = start_date_2.1
  e_d = end_date_2.1
  paste0('https://hydro1.gesdisc.eosdis.nasa.gov/daac-bin/access/timeseries.cgi?variable=GLDAS2:', model, ':', var, '&startDate=', s_d, 'T00:00&endDate=', e_d, 'T21:00&location=GEOM:POINT(', site_info$long, ',%20', site_info$lat, ')&type=asc2')
}

#make a list of all wwws for download - 2.1 only
v2.1_www = unlist(map2(rep(gldas_mod, times = length(params)), params, make_www_2.1))

# for loop to download data - v 2.1 only
for(w21 in 1:length(v2.1_www)) {
  download.file(url = v2.1_www[w21], destfile = file.path(gldas_dir, paste0(rep(gldas_mod, length(params))[w21], '_', params[w21], '.csv')))
}

#make list of files fo GLDAS data
files = list.files(gldas_dir)

formatGLDAS = function(file = files[1]){
  f = read.delim(file.path(gldas_dir, file), skip = 12, sep = '\t')
  colnames(f) = 'data'
  f = f %>%
    rownames_to_column('datetime') %>% 
    mutate(data = as.character(data),
           datetime = as.character(datetime),
           parameter = unlist(str_split(file, pattern = '_'))[5],
           version = unlist(str_split(file, pattern = '_'))[4])
  return(f)
}

# read in downloaded data
all_gldas <- map_dfr(files, formatGLDAS) %>% 
  mutate(datetime_gmt = as.POSIXct(datetime, tz = 'Etc/GMT+0'),
         data = as.numeric(data)) %>%
  arrange(datetime_gmt)

# pivot downloaded data to horizontal
all_gldas_h <- all_gldas %>% 
  group_by(datetime, parameter) %>% 
  summarise(aggrate_data = mean(data)) %>% 
  pivot_wider(names_from = c('parameter'),
              values_from = 'aggrate_data')

# correct time and combine to daily data
all_gldas_h$datetime_gmt <- as.POSIXct(all_gldas_h$datetime, 
                                      tz = 'Etc/GMT+0')#all GLDAS is in GMT

all_gldas_h$datetime_local <- with_tz(all_gldas_h$datetime_gmt,
                                     tz = 'Etc/GMT+6') #NOTE TZ IS INTENTIONALLY INVERTED

all_gldas_h$date <- as.Date(all_gldas_h$datetime_local)

#summarize to daily data - limited to rain and wind data I loaded
gldas_daily <- all_gldas_h %>% 
  group_by(date) %>% 
  summarise(precip = sum(Rainf),
            max_wind = max(Wind),
            mean_wind = mean(Wind),
            min_wind = min(Wind)) %>% 
  rowid_to_column() %>%
  arrange(date)


# the data is downloaded and saved as a dataframe and can be visualized as needed.

# here's a basic plot of max wind by day, idk why but it's getting windier
gldas_daily %>% 
  ggplot(aes(x = date, y = max_wind)) +
  geom_point()

```



## Watershed averaged climate data

Using climate engine, download the two same parameters but for the watershed.
The watershed is stored as a .geojson in the yampa folder. Note, you likely need
to convert the watershed to a ".shp" file before getting the data from climate 
engine. 
#### the watershed stored in the yampa folder is a streamstats output, so it has two shapefiles: a watershed outline and pt. 
#### grab only the watershed outline, not the point.

```{r}
# read in geojson, and look at structure with str() in console
yampa_raw <- st_read("data/yampa/yampa.geojson")

# remove point shapefile
yampa_watershed <- yampa_raw[2,] %>%  # select only the second row
  st_transform(2163)

# convert yampa watershed into a shapefile
st_write(yampa_watershed, "data/yampa/yampa_watershed.shp", append = F)
# append = F bc we want to overwrite any exsiting file. see help page for more info


```


## Compare your climate data anyway you want

Make at least two plots comparing your point versus watershed-averaged climate data. 



