---
title: "2 comparing climate data"
author: "Matthew Ross"
date: "2024-04-17"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(RcppRoll)

gldas_dir = 'data/GLDAS/'

if(!file.exists(gldas_dir)){
  dir.create('data')
  dir.create(gldas_dir)
}
```

# Assignment

For this assignment we are going to compare climate data from a single
point versus a watershed averaged climate data. We will be working over
the Yampa river watershed, one of the last undammed watersheds in the
USA.

## Point-climate acquisition.

Using the 1_climate_data_yojoa.Rmd as an example download at least two
types of climate (wind/temp/etc...) GLDAS data for the Yampa River above
Elkhead Creek. This is the site of a USGS gage that has data dating back
to 2004.

```{r}
site_info <- tibble(site_no = '09244490',
                    lat = 40.5180278,
                    long = -107.3997838,
                    name = 'Yampa_hayden')
```

```{r}
gldas_mod = ('GLDAS_NOAH025_3H_v2.1')
params = c('Tair_f_inst', 'Rainf_f_tavg')

start_date_2.1 = '2004-01-01'
end_date_2.1 = '2023-01-01'

lat = 40.5180278
lon = -107.3997838

make_www_2.1 = function(model, var){#, s_d, e_d, lat, lon) {
  s_d = start_date_2.1
  e_d = end_date_2.1
  paste0('https://hydro1.gesdisc.eosdis.nasa.gov/daac-bin/access/timeseries.cgi?variable=GLDAS2:', model, ':', var, '&startDate=', s_d, 'T00:00&endDate=', e_d, 'T21:00&location=GEOM:POINT(', lon, ',%20', lat, ')&type=asc2')
}

v2.1_www = unlist(map2(rep(gldas_mod, times = length(params)), params, make_www_2.1))

for(w21 in 1:length(v2.1_www)) {
  download.file(url = v2.1_www[w21], destfile = file.path(gldas_dir, paste0(rep(gldas_mod, length(params))[w21], '_', params[w21], '.csv')))
}

#make list of files fo GLDAS data
files = list.files(gldas_dir)


formatGLDAS = function(file = files[1]){
  f = read.delim(file.path(gldas_dir, file), skip = 12, sep = '\t')
  colnames(f) = 'data'
  f = f %>%
    rownames_to_column('datetime') %>% 
    mutate(data = as.character(data),
           datetime = as.character(datetime),
           parameter = unlist(str_split(file, pattern = '_'))[5],
           version = unlist(str_split(file, pattern = '_'))[4])
  return(f)
}

all_gldas = map_dfr(files, formatGLDAS) %>% 
  mutate(datetime_gmt = as.POSIXct(datetime, tz = 'Etc/GMT+0'),
         data = as.numeric(data)) %>%
  arrange(datetime_gmt)

View(all_gldas)

#plot for reality check
ggplot(all_gldas %>%
         sample_frac(0.05), aes(x = datetime_gmt, y = data)) +
  geom_point() +
  facet_grid(parameter ~ ., scales = 'free_y') +
  theme_bw()

all_gldas_h = all_gldas %>% 
  group_by(datetime, parameter) %>% 
  summarise(aggrate_data = mean(data)) %>% 
  pivot_wider(names_from = c('parameter'),
              values_from = 'aggrate_data')

##Summarize data in 5 and 7 days previous

all_gldas_h$datetime_gmt = as.POSIXct(all_gldas_h$datetime, 
                                      tz = 'Etc/GMT+0')#all GLDAS is in GMT

all_gldas_h$datetime_local = with_tz(all_gldas_h$datetime_gmt,
                                     tz = 'Etc/GMT+6') #NOTE TZ IS INTENTIONALLY INVERTED

all_gldas_h$date = as.Date(all_gldas_h$datetime_local)

#summarize to daily data
gldas_daily = all_gldas_h %>% 
  group_by(date) %>% 
  summarise(max_temp = max(Tair),
            min_temp = min(Tair),
            precip = sum(Rainf)) %>% 
  rowid_to_column() %>% 
  filter(date >= as.Date('1980-01-01')) %>% 
  arrange(date) 

#Rolling average for 7 day window prior to date
sevenday = as.data.frame(gldas_daily$date[7:nrow(gldas_daily)])
colnames(sevenday) = 'date'
sevenday$max_temp_7 = roll_max(x  = gldas_daily$max_temp, align = 'right', 7)
sevenday$min_temp_7 = roll_min(x  = gldas_daily$min_temp, align = 'right', 7)
sevenday$precip_7 = roll_sum(x  = gldas_daily$precip, align = 'right', 7)

#Rolling average for 5 day window prior to date
fiveday = as.data.frame(gldas_daily$date[5:nrow(gldas_daily)])
colnames(fiveday) = 'date'
fiveday$max_temp_5 = roll_max(x  = gldas_daily$max_temp, align = 'right', 5)
fiveday$min_temp_5 = roll_min(x  = gldas_daily$min_temp, align = 'right', 5)
fiveday$precip_5 = roll_sum(x  = gldas_daily$precip, align = 'right', 5)

#plot some data
ggplot(fiveday, aes(x = max_temp_5,
                    y = precip_5)) + 
  geom_point()


ggplot(fiveday, aes(x = min_temp_5,
                    y = precip_5)) + 
  geom_point()

##Join GLDAS summaries and export

GLDAS_summary = inner_join(fiveday, sevenday)

ggplot(GLDAS_summary, aes(x = max_temp_5,
                          y = max_temp_7)) + 
  geom_point()


write.csv(GLDAS_summary, file.path(gldas_dir, 'GLDAS_summaries.csv'), row.names = F) 
```

## Watershed averaged climate data

Using climate engine, download the two same parameters but for the
watershed. The watershed is stored as a .geojson in the yampa folder.
Note, you likely need to convert the watershed to a ".shp" file before
getting the data from climate engine.

## Compare your climate data anyway you want

Make at least two plots comparing your point versus watershed-averaged
climate data.
